---
title: "Gentrification Analysis"
author: "Ben Herndon-Miller"
date: "4/14/2019"
output: pdf_document
---

# Load software and data

```{r setup, include=FALSE}
# Load packages
library(knitr) 
library(tidyverse)
library(rgdal)
library(tigris)
library(leaflet)
library(mapview)
library(ggplot2)
library(caret)
library(glmnet)
library(randomForest)
library(xgboost)

# Set default knitr options for knitting code into the report:
opts_chunk$set(echo=TRUE,  # change to FALSE to keep code out of the knitted document
               cache=FALSE, # re-run code that has already been run?
               autodep=TRUE, # assure that caching dependencies are updated correctly
               cache.comments=FALSE, # do not re-run a chunk if only comments are changed
               message=FALSE, # change to FALSE to keep messages out of the knitted document
               warning=FALSE,  # change to FALSE to keep warnings out of the knitted document
               comment = NA,
               tidy = TRUE,
               tidy.opts=list(width.cutoff=60))

# Set seed for reproducibility
set.seed(123)
```

## Source data and make map labels

```{r}
# Load gentrification data
source('clean_gentrification_data.R')

# Convert GEOID to character
acs_data$GEOID <- as.character(acs_data$GEOID)

# Load shape data
houston_shps <- tracts("TX", county = c("Harris County", "Montgomery County", "Fort Bend County"))

# Create dataset for map
houston_shps@data <- houston_shps@data %>%
  left_join(acs_data, by = "GEOID") %>%
  mutate(pop_label = ifelse(pop_change >= 0, 
                            paste0("+",round(pop_change*100,1),"%"),
                            paste0(round(pop_change*100,1),"%")),
         inpov_label = ifelse(inpov_change >= 0, 
                              paste0("+",round(inpov_change*100,1),"%"),
                              paste0(round(inpov_change*100,1),"%")),
         nonwhite_label = ifelse(nonwhite_change >= 0, 
                                 paste0("+",round(nonwhite_change*100,1),"%"),
                                 paste0(round(nonwhite_change*100,1),"%")),
         college_label = ifelse(college_change >= 0, 
                                paste0("+",round(college_change*100,1),"%"),
                                paste0(round(college_change*100,1),"%")),
         med_inc_label = ifelse(med_inc_change >= 0, 
                                paste0("+",round(med_inc_change*100,1),"%"),
                                paste0(round(med_inc_change*100,1),"%")),
         med_rent_label = ifelse(med_rent_change >= 0, 
                                 paste0("+",round(med_rent_change*100,1),"%"),
                                 paste0(round(med_rent_change*100,1),"%")),
         med_val_label = ifelse(med_val_change >= 0, 
                                paste0("+",round(med_val_change*100,1),"%"),
                                paste0(round(med_val_change*100,1),"%")),
         total_renters_label = ifelse(renters_change >= 0, 
                                paste0("+",round(renters_change*100,1),"%"),
                                paste0(round(renters_change*100,1),"%")),
         total_houses_label = ifelse(total_houses_change >= 0, 
                                     paste0("+",round(total_houses_change*100,1),"%"),
                                     paste0(round(total_houses_change*100,1),"%")))
```

# Visualizations

## Opportunity Zone Map

```{r}
# Define logo for map 
img <- "http://januaryadvisors.com/wp-content/uploads/2013/05/ja-og.png"

# Subset data frames for plotting
ozs <- subset(houston_shps, op_zone == 1)
eligible_ozs <- subset(houston_shps, op_zone == 0 & op_zone_eligible == 1)

# Create labels for the map
tract_labels <- sprintf(
  paste0("<strong>GEOID: %s</strong>",
         "<br/>Total Population: %s",
         "<br/>Non-white Population: %s",
         "<br/>Low-income Population: %s",
         "<br/>College Educated Population: %s",
         "<br/>Median Income: %s",
         "<br/>Median Rent: %s",
         "<br/>Median Home Value: %s",
         "<br/>Number of Renters: %s",
         "<br/>Number of Houses: %s"),
  houston_shps@data$GEOID, 
  houston_shps@data$pop_label, 
  houston_shps@data$nonwhite_label,
  houston_shps@data$inpov_label,
  houston_shps@data$college_label,
  houston_shps@data$med_inc_label,
  houston_shps@data$med_rent_label,
  houston_shps@data$med_val_label,
  houston_shps@data$total_renters_label,
  houston_shps@data$total_houses_label) %>% 
  lapply(htmltools::HTML)

oz_tract_labels <- sprintf(
  paste0("<strong>GEOID: %s</strong>",
         "<br/>Total Population: %s",
         "<br/>Non-white Population: %s",
         "<br/>Low-income Population: %s",
         "<br/>College Educated Population: %s",
         "<br/>Median Income: %s",
         "<br/>Median Rent: %s",
         "<br/>Median Home Value: %s",
         "<br/>Number of Renters: %s",
         "<br/>Number of Houses: %s"),
  ozs@data$GEOID, 
  ozs@data$pop_label, 
  ozs@data$nonwhite_label,
  ozs@data$inpov_label,
  ozs@data$college_label,
  ozs@data$med_inc_label,
  ozs@data$med_rent_label,
  ozs@data$med_val_label,
  ozs@data$total_renters_label,
  ozs@data$total_houses_label) %>% 
  lapply(htmltools::HTML)

eligible_tract_labels <- sprintf(
  paste0("<strong>GEOID: %s</strong>",
         "<br/>Total Population: %s",
         "<br/>Non-white Population: %s",
         "<br/>Low-income Population: %s",
         "<br/>College Educated Population: %s",
         "<br/>Median Income: %s",
         "<br/>Median Rent: %s",
         "<br/>Median Home Value: %s",
         "<br/>Number of Renters: %s",
         "<br/>Number of Houses: %s"),
  eligible_ozs@data$GEOID, 
  eligible_ozs@data$pop_label, 
  eligible_ozs@data$nonwhite_label,
  eligible_ozs@data$inpov_label,
  eligible_ozs@data$college_label,
  eligible_ozs@data$med_inc_label,
  eligible_ozs@data$med_rent_label,
  eligible_ozs@data$med_val_label,
  eligible_ozs@data$total_renters_label,
  eligible_ozs@data$total_houses_label) %>% 
  lapply(htmltools::HTML)

# Map of Opportunity Zones
oz_map <- leaflet() %>%
      addTiles() %>% 
      setView(-95.37, 29.76, zoom = 9) %>%
      addPolygons(data = houston_shps,
          fillColor = "#000000",
          weight = 1,
          opacity = 0.7,
          color = "#FFFFFF",
          dashArray = "",
          fillOpacity = 0,
          highlight = highlightOptions(
            weight = 3,
            color = "#000000",
            dashArray = "",
            fillOpacity = 0,
            bringToFront = TRUE),
          label = tract_labels,
          labelOptions = labelOptions(
            style = list("font-weight" = "normal", padding = "3px 8px"),
            textsize = "12px",
            direction = "auto")) %>%
      addPolygons(data = ozs,
              fillColor = "#F5811D",
              weight = 1,
              opacity = 0.7,
              color = "#FFFFFF",
              dashArray = "",
              fillOpacity = 0.7,
              group = "Opportunity Zone",
              highlight = highlightOptions(
                weight = 3,
                color = "#000000",
                dashArray = "",
                fillOpacity = 0.7,
                bringToFront = TRUE),
              label = oz_tract_labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "12px",
                direction = "auto")) %>%
        addPolygons(data = eligible_ozs,
              fillColor = "#4776BA",
              weight = 1,
              opacity = 0.7,
              color = "#FFFFFF",
              dashArray = "",
              fillOpacity = 0.7,
              group = "Eligible Opportunity Zone",
              highlight = highlightOptions(
                weight = 3,
                color = "#000000",
                dashArray = "",
                fillOpacity = 0.7,
                bringToFront = TRUE),
              label = eligible_tract_labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "12px",
                direction = "auto")) %>%
        addLayersControl(overlayGroups = c("Opportunity Zone", "Eligible Opportunity Zone")) %>%
        hideGroup("Eligible Opportunity Zone") %>%
        addLogo(img, url = "http://januaryadvisors.com/wp-content/uploads/2013/05/ja-og", 
              position = "bottomleft", height = 100, width = 100, offset.x = 20, offset.y = 20) %>%
        addLegend(position = "bottomright", colors = c("#F5811D", "#4776BA"), 
                  labels = c("Opportunity Zones", "Eligible Opportunity Zones"), opacity = 0.7)

oz_map
```

## Gentrification Map

```{r}
# Subset data frames for plotting
vulnerable <- subset(houston_shps, vulnerable == 1)
dem_gent <- subset(houston_shps, dem_gentrification == 1)
inv_gent <- subset(houston_shps, inv_gentrification == 1)
fed_vuln <- subset(houston_shps, fed_vulnerable == 1)
fed_gent <- subset(houston_shps, fed_gentrifying == 1)

# Create labels for map
vulnerable_tract_labels <- sprintf(
  paste0("<strong>GEOID: %s</strong>",
         "<br/>Total Population: %s",
         "<br/>Non-white Population: %s",
         "<br/>Low-income Population: %s",
         "<br/>College Educated Population: %s",
         "<br/>Median Income: %s",
         "<br/>Median Rent: %s",
         "<br/>Median Home Value: %s",
         "<br/>Number of Renters: %s",
         "<br/>Number of Houses: %s"),
  vulnerable@data$GEOID, 
  vulnerable@data$pop_label, 
  vulnerable@data$nonwhite_label,
  vulnerable@data$inpov_label,
  vulnerable@data$college_label,
  vulnerable@data$med_inc_label,
  vulnerable@data$med_rent_label,
  vulnerable@data$med_val_label,
  vulnerable@data$total_renters_label,
  vulnerable@data$total_houses_label) %>% 
  lapply(htmltools::HTML)

dem_gent_tract_labels <- sprintf(
  paste0("<strong>GEOID: %s</strong>",
         "<br/>Total Population: %s",
         "<br/>Non-white Population: %s",
         "<br/>Low-income Population: %s",
         "<br/>College Educated Population: %s",
         "<br/>Median Income: %s",
         "<br/>Median Rent: %s",
         "<br/>Median Home Value: %s",
         "<br/>Number of Renters: %s",
         "<br/>Number of Houses: %s"),
  dem_gent@data$GEOID, 
  dem_gent@data$pop_label, 
  dem_gent@data$nonwhite_label,
  dem_gent@data$inpov_label,
  dem_gent@data$college_label,
  dem_gent@data$med_inc_label,
  dem_gent@data$med_rent_label,
  dem_gent@data$med_val_label,
  dem_gent@data$total_renters_label,
  dem_gent@data$total_houses_label) %>% 
  lapply(htmltools::HTML)

inv_gent_tract_labels <- sprintf(
  paste0("<strong>GEOID: %s</strong>",
         "<br/>Total Population: %s",
         "<br/>Non-white Population: %s",
         "<br/>Low-income Population: %s",
         "<br/>College Educated Population: %s",
         "<br/>Median Income: %s",
         "<br/>Median Rent: %s",
         "<br/>Median Home Value: %s",
         "<br/>Number of Renters: %s",
         "<br/>Number of Houses: %s"),
  inv_gent@data$GEOID, 
  inv_gent@data$pop_label, 
  inv_gent@data$nonwhite_label,
  inv_gent@data$inpov_label,
  inv_gent@data$college_label,
  inv_gent@data$med_inc_label,
  inv_gent@data$med_rent_label,
  inv_gent@data$med_val_label,
  inv_gent@data$total_renters_label,
  inv_gent@data$total_houses_label) %>% 
  lapply(htmltools::HTML)

fed_vuln_tract_labels <- sprintf(
  paste0("<strong>GEOID: %s</strong>",
         "<br/>Total Population: %s",
         "<br/>Non-white Population: %s",
         "<br/>Low-income Population: %s",
         "<br/>College Educated Population: %s",
         "<br/>Median Income: %s",
         "<br/>Median Rent: %s",
         "<br/>Median Home Value: %s",
         "<br/>Number of Renters: %s",
         "<br/>Number of Houses: %s"),
  fed_vuln@data$GEOID, 
  fed_vuln@data$pop_label, 
  fed_vuln@data$nonwhite_label,
  fed_vuln@data$inpov_label,
  fed_vuln@data$college_label,
  fed_vuln@data$med_inc_label,
  fed_vuln@data$med_rent_label,
  fed_vuln@data$med_val_label,
  fed_vuln@data$total_renters_label,
  fed_vuln@data$total_houses_label) %>% 
  lapply(htmltools::HTML)

fed_gent_tract_labels <- sprintf(
  paste0("<strong>GEOID: %s</strong>",
         "<br/>Total Population: %s",
         "<br/>Non-white Population: %s",
         "<br/>Low-income Population: %s",
         "<br/>College Educated Population: %s",
         "<br/>Median Income: %s",
         "<br/>Median Rent: %s",
         "<br/>Median Home Value: %s",
         "<br/>Number of Renters: %s",
         "<br/>Number of Houses: %s"),
  fed_gent@data$GEOID, 
  fed_gent@data$pop_label, 
  fed_gent@data$nonwhite_label,
  fed_gent@data$inpov_label,
  fed_gent@data$college_label,
  fed_gent@data$med_inc_label,
  fed_gent@data$med_rent_label,
  fed_gent@data$med_val_label,
  fed_gent@data$total_renters_label,
  fed_gent@data$total_houses_label) %>% 
  lapply(htmltools::HTML)

# Map of Gentrified Census Tracts
gent_map <- leaflet() %>%
      addTiles() %>% 
      setView(-95.37, 29.76, zoom = 9) %>%
      addPolygons(data = houston_shps,
          fillColor = "#000000",
          weight = 1,
          opacity = 0.7,
          color = "#FFFFFF",
          dashArray = "",
          fillOpacity = 0,
          highlight = highlightOptions(
            weight = 3,
            color = "#000000",
            dashArray = "",
            fillOpacity = 0,
            bringToFront = TRUE),
          label = tract_labels,
          labelOptions = labelOptions(
            style = list("font-weight" = "normal", padding = "3px 8px"),
            textsize = "12px",
            direction = "auto")) %>%
      addPolygons(data = vulnerable,
          fillColor = "#13C18D",
          weight = 1,
          opacity = 0.7,
          color = "#FFFFFF",
          dashArray = "",
          fillOpacity = 0.7,
          group = "Kinder: Gentrifiable",
          highlight = highlightOptions(
            weight = 3,
            color = "#000000",
            dashArray = "",
            fillOpacity = 0.7,
            bringToFront = TRUE),
          label = vulnerable_tract_labels,
          labelOptions = labelOptions(
            style = list("font-weight" = "normal", padding = "3px 8px"),
            textsize = "12px",
            direction = "auto")) %>%
      addPolygons(data = dem_gent,
          fillColor = "#A71183",
          weight = 1,
          opacity = 0.7,
          color = "#FFFFFF",
          dashArray = "",
          fillOpacity = 0.7,
          group = "Kinder: Gentrifying Demographically",
          highlight = highlightOptions(
            weight = 3,
            color = "#000000",
            dashArray = "",
            fillOpacity = 0.7,
            bringToFront = TRUE),
          label = dem_gent_tract_labels,
          labelOptions = labelOptions(
            style = list("font-weight" = "normal", padding = "3px 8px"),
            textsize = "12px",
            direction = "auto")) %>%
      addPolygons(data = inv_gent,
          fillColor = "#2575B7",
          weight = 1,
          opacity = 0.7,
          color = "#FFFFFF",
          dashArray = "",
          fillOpacity = 0.7,
          group = "Kinder: Gentrifying Financially",
          highlight = highlightOptions(
            weight = 3,
            color = "#000000",
            dashArray = "",
            fillOpacity = 0.7,
            bringToFront = TRUE),
          label = inv_gent_tract_labels,
          labelOptions = labelOptions(
            style = list("font-weight" = "normal", padding = "3px 8px"),
            textsize = "12px",
            direction = "auto")) %>%
      addPolygons(data = fed_vuln,
          fillColor = "#C11344",
          weight = 1,
          opacity = 0.7,
          color = "#FFFFFF",
          dashArray = "",
          fillOpacity = 0.7,
          group = "Federal Reserve: Gentrifiable",
          highlight = highlightOptions(
            weight = 3,
            color = "#000000",
            dashArray = "",
            fillOpacity = 0.7,
            bringToFront = TRUE),
          label = fed_vuln_tract_labels,
          labelOptions = labelOptions(
            style = list("font-weight" = "normal", padding = "3px 8px"),
            textsize = "12px",
            direction = "auto")) %>%
      addPolygons(data = fed_gent,
          fillColor = "#081F30",
          weight = 1,
          opacity = 0.7,
          color = "#FFFFFF",
          dashArray = "",
          fillOpacity = 0.9,
          group = "Federal Reserve: Gentrifying",
          highlight = highlightOptions(
            weight = 3,
            color = "#000000",
            dashArray = "",
            fillOpacity = 0.9,
            bringToFront = TRUE),
          label = fed_gent_tract_labels,
          labelOptions = labelOptions(
            style = list("font-weight" = "normal", padding = "3px 8px"),
            textsize = "12px",
            direction = "auto")) %>%
      addLayersControl(baseGroups = c("Federal Reserve: Gentrifiable", "Federal Reserve: Gentrifying", "Kinder: Gentrifiable", "Kinder: Gentrifying Demographically", "Kinder: Gentrifying Financially")) %>%
      addLogo(img, url = "http://januaryadvisors.com/wp-content/uploads/2013/05/ja-og", 
              position = "bottomleft", height = 100, width = 100, offset.x = 20, offset.y = 20) %>%
      addLegend(position = "bottomright", colors = c("#C11344", "#081F30", "#13C18D", "#A71183", "#2575B7"), 
                  labels = c("Federal Reserve: Gentrifiable", "Federal Reserve: Gentrifying", "Kinder: Gentrifiable", "Kinder: Gentrifying Demographically", "Kinder: Gentrifying Financially"), opacity = 0.7)

gent_map
```

## Gentrification Risk-Factors by Op-Zone Status

### Wrangle data

```{r}
# Create factor level of census tract status for grouping
acs_data$tract_status <- NA
acs_data$tract_status[acs_data$op_zone == 1] <- "Opportunity Zone"
acs_data$tract_status[acs_data$op_zone == 0 & acs_data$op_zone_eligible == 1] <- "Eligible"
acs_data$tract_status[acs_data$op_zone == 0 & acs_data$op_zone_eligible == 0] <- "Not-Eligible"

# Calculate gentrification status by tract status
gent_data <- acs_data %>%
  filter(!is.na(tract_status)) %>%
  group_by(tract_status) %>%
  summarise(vulnerability_avg = mean(vulnerable, na.rm = TRUE),
            dem_change_avg = mean(dem_gentrification, na.rm = TRUE),
            inv_change_avg = mean(inv_gentrification, na.rm = TRUE),
            fed_vuln_avg = mean(fed_vulnerable, na.rm = TRUE),
            fed_gent_avg = mean(fed_gentrifying, na.rm = TRUE)) %>%
  gather("variable", "value", -tract_status) %>%
  mutate(tract_order = case_when(tract_status == "Opportunity Zone" ~ "C",
                                 tract_status == "Eligible" ~ "B",
                                 tract_status == "Not-Eligible" ~ "A"),
         var_order = case_when(variable == "vulnerability_avg" ~ "D",
                               variable == "fed_vuln_avg" ~ "E",
                               variable == "fed_gent_avg" ~ "C",
                               variable == "dem_change_avg" ~ "B",
                               variable == "inv_change_avg" ~ "A"))

# Set nonwhite quantiles
nonwhite_quantiles <- seq(0,1,0.1)

# Calculate op zone status by nonwhite status
nonwhite_ozs <- acs_data %>%
  mutate(non_white_bin = .bincode(pct_nonwhite_2016, nonwhite_quantiles, TRUE, TRUE)) %>%
  group_by(non_white_bin) %>%
  summarise(pct_opzones = mean(op_zone, na.rm = TRUE),
            tracts = n())

# Set college quantiles
college_quantiles <- seq(0,0.7,0.1)

# Calculate op zone status by college status
college_ozs <- acs_data %>%
  mutate(college_bin = .bincode(pct_college_2016, college_quantiles, TRUE, TRUE)) %>%
  group_by(college_bin) %>%
  summarise(pct_opzones = mean(op_zone, na.rm = TRUE),
            tracts = n())

# Set renter quantiles
renter_quantiles <- seq(0,1,0.1)

# Calculate op zone status by renter status
renter_ozs <- acs_data %>%
  mutate(renter_bin = .bincode(pct_renters_2016, renter_quantiles, TRUE, TRUE)) %>%
  group_by(renter_bin) %>%
  summarise(pct_opzones = mean(op_zone, na.rm = TRUE),
            tracts = n())

# Set in-poverty quantiles
inpov_quantiles <- c(seq(0,0.4,0.1),1)

# Calculate op zone status by in-poverty status
inpov_ozs <- acs_data %>%
  mutate(inpov_bin = .bincode(pct_inpov_2016, inpov_quantiles, TRUE, TRUE)) %>%
  group_by(inpov_bin) %>%
  summarise(pct_opzones = mean(op_zone, na.rm = TRUE),
            tracts = n())
  
```

### Bar plot by gentrification criteria

```{r}
# Grouped bar plot
gent_plot <- ggplot(gent_data, aes(var_order, value)) +
  geom_bar(aes(fill = tract_order), position = "dodge", stat = "identity") + 
  scale_fill_manual(labels = c("Not Eligible", "Eligible", "Opportunity Zone"), 
                    values = c("#cccccc", "#319CF4", "#FF8B00")) +
  scale_x_discrete(labels = c("Kinder: Gentrifying Financially", "Kinder: Gentrifying Demographically","Federal Reserve: Gentrifying", "Kinder: Gentrifiable", "Federal Reserve: Gentrifiable")) +
  scale_y_continuous(breaks = seq(0,1,0.1), 
                     labels = paste0(seq(0,1,0.1)*100,"%")) +
  ylab("% Meeting Criteria") +
  labs(fill = "Census Tract Designation") +
  #ggtitle("Gentrification Criteria by Opportunity Zone Status") +
  theme_light() + 
  theme(plot.title = element_text(family = "Roboto", size = 20),
        axis.title.x = element_text(family = "Roboto", size = 20, vjust = -5),
        axis.title.y = element_blank(),
        legend.title = element_text(family = "Roboto", size = 14),
        legend.text = element_text(family = "Roboto", size = 14),
        axis.text.x = element_text(family = "Roboto", size = 14),
        axis.text.y = element_text(family = "Roboto", size = 14),
        panel.border = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin=unit(c(1,1,1.5,1.2),"cm")) +
  guides(fill = guide_legend(reverse=TRUE)) +
  coord_flip()

gent_plot
```

### Bar plot by nonwhite population

```{r}
# Grouped bar plot
nonwhite_plot <- ggplot(nonwhite_ozs, aes(non_white_bin, pct_opzones)) +
  geom_bar(stat = "identity", fill = "#FF8B00") +
  scale_x_continuous(breaks = seq(1,10),
                     labels = paste0(seq(0.1,1,0.1)*100,"%")) + 
  scale_y_continuous(breaks = seq(0,0.8,0.1), 
                     labels = paste0(seq(0,0.8,0.1)*100,"%")) +
  ylab("Census Tracts Designated Opportunity Zones") + 
  xlab("Non-white Population") + 
  theme_light() + 
  theme(plot.title = element_text(family = "Roboto", size = 20),
        axis.title.x = element_text(family = "Roboto", size = 20, vjust = -4),
        axis.title.y = element_text(family = "Roboto", size = 20,  vjust = 4),
        legend.title = element_text(family = "Roboto"),
        legend.text = element_text(family = "Roboto"),
        axis.text.x = element_text(family = "Roboto", size = 16),
        axis.text.y = element_text(family = "Roboto", size = 16),
        panel.border = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin=unit(c(1,1,1.5,1.2),"cm"))

nonwhite_plot
```


### Bar plot by college educated population

```{r}
# Grouped bar plot
college_plot <- ggplot(college_ozs, aes(college_bin, pct_opzones)) +
  geom_bar(stat = "identity", fill = "#319CF4") +
  scale_x_continuous(breaks = seq(1,7),
                     labels = paste0(seq(0.1,0.7,0.1)*100,"%")) + 
  scale_y_continuous(breaks = seq(0,0.3,0.05), 
                     labels = paste0(seq(0,0.3,0.05)*100,"%")) +
  ylab("% of Census Tracts") + 
  xlab("College Educated Population") + 
  theme_light() + 
  theme(plot.title = element_text(family = "Roboto", size = 20),
        axis.title.x = element_text(family = "Roboto", size = 20, vjust = -4),
        axis.title.y = element_text(family = "Roboto", size = 20,  vjust = 4),
        legend.title = element_text(family = "Roboto"),
        legend.text = element_text(family = "Roboto"),
        axis.text.x = element_text(family = "Roboto", size = 16),
        axis.text.y = element_text(family = "Roboto", size = 16),
        panel.border = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin=unit(c(1,1,1.5,1.2),"cm"))

college_plot
```

### Bar plot by renter population

```{r}
# Grouped bar plot
renter_plot <- ggplot(renter_ozs, aes(renter_bin, pct_opzones)) +
  geom_bar(stat = "identity", fill = "#13C18D") +
  scale_x_continuous(breaks = seq(1,10),
                     labels = paste0(seq(0.1,1,0.1)*100,"%")) + 
  scale_y_continuous(breaks = seq(0,0.4,0.05), 
                     labels = paste0(seq(0,0.4,0.05)*100,"%")) +
  ylab("% of Census Tracts") + 
  xlab("Population Renting") + 
  theme_light() + 
  theme(plot.title = element_text(family = "Roboto", size = 20),
        axis.title.x = element_text(family = "Roboto", size = 20, vjust = -4),
        axis.title.y = element_text(family = "Roboto", size = 20,  vjust = 4),
        legend.title = element_text(family = "Roboto"),
        legend.text = element_text(family = "Roboto"),
        axis.text.x = element_text(family = "Roboto", size = 16),
        axis.text.y = element_text(family = "Roboto", size = 16),
        panel.border = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin=unit(c(1,1,1.5,1.2),"cm"))

renter_plot
```


### Bar plot by in-poverty population

```{r}
# Grouped bar plot
inpov_plot <- ggplot(inpov_ozs, aes(inpov_bin, pct_opzones)) +
  geom_bar(stat = "identity", fill = "#A71183") +
  scale_x_continuous(breaks = seq(1,5),
                     labels = c("10%", "20%", "30%", "40%", ">50%")) + 
  scale_y_continuous(breaks = seq(0,0.5,0.05), 
                     labels = paste0(seq(0,0.5,0.05)*100,"%")) +
  ylab("% of Census Tracts") + 
  xlab("Population Below Poverty-Line") + 
  theme_light() + 
  theme(plot.title = element_text(family = "Roboto", size = 20),
        axis.title.x = element_text(family = "Roboto", size = 20, vjust = -4),
        axis.title.y = element_text(family = "Roboto", size = 20,  vjust = 4),
        legend.title = element_text(family = "Roboto"),
        legend.text = element_text(family = "Roboto"),
        axis.text.x = element_text(family = "Roboto", size = 16),
        axis.text.y = element_text(family = "Roboto", size = 16),
        panel.border = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin=unit(c(1,1,1.5,1.2),"cm"))

inpov_plot
```

# Modeling

## Pre-Process Data

```{r}
# Pull data for model, choose weakly gentrified to be outcome of interest
acs_model_data <- acs_data %>%
  select(c(vulnerable, nonwhite_change, black_change, asian_change, hispanic_change,
           renters_change, inpov_change, college_change, pop_change, med_rent_change,
           med_val_change, total_houses_change, op_zone, op_zone_eligible))

# Filter for rows that are not NA
acs_model_data <- acs_model_data[complete.cases(acs_model_data),]

# Check for predictors with near-zero variance
nzv <- nearZeroVar(acs_model_data, saveMetrics = TRUE)

# Check for highly correlated variables
cor_vars <- cor(acs_model_data)
high_cor <- sum(abs(cor_vars[upper.tri(cor_vars)]) > .8)

# Check for linear combinations
lin_combos <- findLinearCombos(acs_model_data)

# Change categorical variables to factor
acs_model_data$vulnerable <- as.factor(acs_model_data$vulnerable)
acs_model_data$op_zone <- as.factor(acs_model_data$op_zone)
acs_model_data$op_zone_eligible <- as.factor(acs_model_data$op_zone_eligible)

# Create vector of numeric variables
num_vecs <- unlist(lapply(acs_model_data, is.numeric))

# Center and scale data for prediction
#process_var <- preProcess(acs_model_data[,num_vecs], method = c("center", "scale"))
#acs_model_data[,num_vecs] <- predict(process_var, acs_model_data[,num_vecs])

# Chunk data 
idx <- sample(1:nrow(acs_model_data), nrow(acs_model_data)*.7, replace = FALSE)
train_data <- acs_model_data[idx,]
test_data <- acs_model_data[-idx,]
```

## Fit Logistic Regression Model

```{r}
# Fit logistic regression model
logistic_fit <- glm(data = acs_model_data, formula = vulnerable ~ ., family = "binomial")
summary(logistic_fit)
```

### Visualize Logistic Regression Results

```{r}
log_results <- data.frame(var = names(logistic_fit$coefficients[-1]), val = logistic_fit$coefficients[-1])
```


## Predict Gentrification with Lasso

```{r}
# Create vector of lambdas to tune across
lambdas <- 10^seq(-4, 4, by = .1)

# Perform cross-validation to tune lambda
lasso_cv <- cv.glmnet(data.matrix(train_data[,-1]), train_data[,1], family = "binomial", 
                      alpha = 1, type.measure = "auc", lambda = lambdas)

# Predict on test data
lasso_y_hat <- round(predict(lasso_cv, newx = data.matrix(test_data[,-1]), s = "lambda.min", type = "response"))

# Prediction accuracy
lasso_acc <- mean(lasso_y_hat == test_data[,1])
```

## Selected features

### Non-zero coefficients

```{r}
# Show selected features from optimal lasso
coef(lasso_cv, s = "lambda.min")
```

### Plot coefficient vs. lambda

```{r}
# Function for labels
lbs_fun <- function(fit, ...) {
        L <- length(lasso_fit$lambda)
        x <- log(lasso_fit$lambda[L])
        y <- lasso_fit$beta[, L]
        labs <- names(y)
        text(x, y, labels=labs)
}

# Fit lasso model
lasso_fit <- glmnet(data.matrix(train_data[,-1]), train_data[,1], family = "binomial", 
                      alpha = 1, lambda = lambdas)

# Plot lasso coefficients vs. lambda
plot(lasso_fit, xvar = "lambda", xlim = c(-15, 0))
lbs_fun(lasso_fit)
```

## Predict Gentrification with Ridge

```{r}
# Perform cross-validation to tune lambda
ridge_cv <- cv.glmnet(data.matrix(train_data[,-1]), train_data[,1], family = "binomial", 
                      alpha = 0, type.measure = "auc", lambda = lambdas)

# Predict on test data
ridge_y_hat <- round(predict(ridge_cv, newx = data.matrix(test_data[,-1]), s = "lambda.min", type = "response"))

# Prediction accuracy
ridge_acc <- mean(ridge_y_hat == test_data[,1])
```

## Predict with Random Forest

### Choose number of trees

```{r}
# Run Random Forest model without tuning
tree_tune <- randomForest(formula = vulnerable ~ ., data = train_data)

# Build data frame for errors to plot
tree_tune_err <- data.frame(ntree = 1:length(tree_tune$err.rate[,1]), oob_error = tree_tune$err.rate[,1])

# Plot error 
ggplot(data = tree_tune_err, aes(x = ntree, y = oob_error)) +
  geom_line()
```

### Tune other hyper-parameters

```{r}
# Create hyper-parameter grid
tune_grid <- expand.grid(
  mtry       = seq(1, 13),
  sampsize = round(c(.55, .632, .70, .75, .80)*nrow(train_data)),
  nodesize  = seq(1, 12)
)

# Tune model using OOB error
for(i in 1:nrow(tune_grid)) {
  
  # Train model
  grid_fit <- randomForest(
    formula = vulnerable ~ ., 
    data = train_data, 
    num.trees = 300,
    mtry = tune_grid$mtry[i],
    sampsize = tune_grid$sampsize[i],
    nodesize = tune_grid$nodesize[i])
  
  # Record OOB error
  tune_grid$OOB_error[i] <- grid_fit$err.rate[,1]
}

# Print best combination of results
head(tune_grid[order(tune_grid$OOB_error),])
```

### Predict on test set

```{r}
# Retrain random forest model on optimal hyper-parameter values
rf_best_fit <- randomForest(
    formula = vulnerable ~ ., 
    data = train_data, 
    num.trees = 300,
    mtry = 13,
    sampsize = 428,
    nodesize = 2)

# Predict on test set
rf_y_hat <- predict(rf_best_fit, test_data)

# Prediction accuracy
rf_acc <- mean(rf_y_hat == test_data[,1])
```

## Predict with Xgboost

```{r}
# Change training labels
train_data$vulnerable <- ifelse(train_data$vulnerable == 1, "Yes", "No")
test_data$vulnerable <- ifelse(test_data$vulnerable == 1, "Yes", "No")


# Create tuning controls for 5-fold CV
tune_control <- trainControl(method = "cv", 
                             number = 5,
                             classProbs = TRUE,
                             summaryFunction = twoClassSummary)

# Create tuning grid
tune_grid <- expand.grid(
  nrounds = seq(50, 300, 50),
  max_depth = seq(2,6,2),
  eta = seq(0.1,0.9,0.2),
  gamma = c(0,1,5,10),
  colsample_bytree = seq(0.1,1,.3),
  min_child_weight = c(0,1,5,10),
  subsample = seq(0.1,1,.3)
  )

# Fit model with 5-fold CV
xgb_tune_cv <- train(vulnerable ~ ., data = train_data, 
                 method = "xgbTree", 
                 trControl = tune_control, 
                 verbose = TRUE, 
                 tuneGrid = tune_grid,
                 metric = "ROC")
```

```{r}
# Re-factor training labels
train_data$vulnerable <- as.numeric(ifelse(train_data$vulnerable == "Yes", 1, 0))
test_data$vulnerable <- as.numeric(ifelse(test_data$vulnerable == "Yes", 1, 0))


# Re-fit model at optimal training values
xgb_fit <- xgboost(data = data.matrix(train_data[,-1]),
                   label = train_data[,1],
                   nrounds = xgb_tune_cv$bestTune[["nrounds"]],
                   objective = "binary:logistic",
                   verbose = 0,
                   params = list(max_depth = xgb_tune_cv$bestTune[["max_depth"]],
                                 eta = xgb_tune_cv$bestTune[["eta"]],
                                 gamma = xgb_tune_cv$bestTune[["gamma"]],
                                 colsample_bytree = xgb_tune_cv$bestTune[["colsample_bytree"]],
                                 min_child_weight = xgb_tune_cv$bestTune[["min_child_weight"]],
                                 subsample = xgb_tune_cv$bestTune[["subsample"]]))

# Predict on the test set
xgb_y_hat <- round(predict(xgb_fit, data.matrix(test_data[,-1]), type = "response"))

# Prediction accuracy
xgb_acc <- mean(xgb_y_hat == test_data[,1])
```
